---
output:
    html_document:
        toc: true
        toc_depth: 6
version: 0.2
author: Tamas Ryszard Sztanka-Toth, Nikolaos Karaiskos
email: tamasryszard.sztanka-toth@mdc-berlin.de, nikolaos.karaiskos@mdc.berlin.de
license: GPL
---

```{r knitr_options, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(
  cache = F,
  autodep = TRUE,
  message = FALSE,
  warning = FALSE,
  comment = NA
)
```

```{r load_libraries, echo =F}
library(tidyverse)
library(yaml)
library(cowplot)
library(knitr)
library(grid)
library(gtable)

theme_set(theme_cowplot(12))
```

```{r functions, echo = F}
readStarLog <- function(log_file){
		out = list()
		lines = readLines(log_file)
	
		out$input_reads = (lines[6] %>% strsplit('\t') %>% unlist)[2] %>% as.integer

		out$uniq_mapped_reads = (lines[9] %>% strsplit('\t') %>% unlist)[2] %>% as.integer

		#out$avg_length = (lines[11] %>% strsplit('\t') %>% unlist)[2] %>% as.numeric
		
        tibble(observation=names(out), value=unlist(unname(out)))
	}

shift_legend <- function(p){

  # check if p is a valid object
  if(!"gtable" %in% class(p)){
    if("ggplot" %in% class(p)){
      gp <- ggplotGrob(p) # convert to grob
    } else {
      message("This is neither a ggplot object nor a grob generated from ggplotGrob. Returning original plot.")
      return(p)
    }
  } else {
    gp <- p
  }

  # check for unfilled facet panels
  facet.panels <- grep("^panel", gp[["layout"]][["name"]])
  empty.facet.panels <- sapply(facet.panels, function(i) "zeroGrob" %in% class(gp[["grobs"]][[i]]))
  empty.facet.panels <- facet.panels[empty.facet.panels]
  if(length(empty.facet.panels) == 0){
    message("There are no unfilled facet panels to shift legend into. Returning original plot.")
    return(p)
  }

  # establish extent of unfilled facet panels (including any axis cells in between)
  empty.facet.panels <- gp[["layout"]][empty.facet.panels, ]
  empty.facet.panels <- list(min(empty.facet.panels[["t"]]), min(empty.facet.panels[["l"]]),
                             max(empty.facet.panels[["b"]]), max(empty.facet.panels[["r"]]))
  names(empty.facet.panels) <- c("t", "l", "b", "r")

  # extract legend & copy over to location of unfilled facet panels
  guide.grob <- which(gp[["layout"]][["name"]] == "guide-box")
  if(length(guide.grob) == 0){
    message("There is no legend present. Returning original plot.")
    return(p)
  }
  gp <- gtable_add_grob(x = gp,
                        grobs = gp[["grobs"]][[guide.grob]],
                        t = empty.facet.panels[["t"]],
                        l = empty.facet.panels[["l"]],
                        b = empty.facet.panels[["b"]],
                        r = empty.facet.panels[["r"]],
                        name = "new-guide-box")

  # squash the original guide box's row / column (whichever applicable)
  # & empty its cell
  guide.grob <- gp[["layout"]][guide.grob, ]
  if(guide.grob[["l"]] == guide.grob[["r"]]){
    gp <- gtable_squash_cols(gp, cols = guide.grob[["l"]])
  }
  if(guide.grob[["t"]] == guide.grob[["b"]]){
    gp <- gtable_squash_rows(gp, rows = guide.grob[["t"]])
  }
  gp <- gtable_remove_grobs(gp, "guide-box")

  return(gp)
}
```


```{r read_qc_params_file, echo = F}
qc_params <- read_yaml(snakemake@input[['parameters_file']])
```

#### Run information

- **Project**: `r qc_params$project_id`
- **Sample**: `r qc_params$sample_id`
- **Puck**: `r qc_params$puck_id`
- **Experiment**: `r qc_params$experiment`
- **Sequencing date**: `r qc_params$sequencing_date`

saturation analysis v.0.1.1, generated on `r format(Sys.time(), '%d/%B/%Y')`

contact: tamasryszard.sztanka-toth@mdc-berlin.de, nikolaos.karaiskos@mdc-berlin.de

#### Downstream stats

```{r parse_metadata, echo = F}
metadata <- readStarLog(snakemake@input$star_log) %>%
    rbind(read_table2(snakemake@input$reads_type_out, col_names=c('observation', 'value'))) %>%
    # convert to millions
    mutate(value = round(value / 1e6, 2)) %>%
    {
        mutate(., input_reads = filter(., observation == 'input_reads')$value)
    } %>%
    mutate(label = ifelse(observation == 'input_reads', value, paste0(value, ' (', round(value/input_reads*100, 1), '%)'))) %>%
    dplyr::select(observation, label) %>%
    spread(observation, label) %>%
    dplyr::rename(as.utr = UTR,
                  intronic = INTRONIC,
                  intergenic = INTERGENIC,
                  as.cds = CODING) %>%
    # reorder columns
    dplyr::select(input_reads, uniq_mapped_reads, intergenic, intronic, as.cds, as.utr) %>%
    kable

metadata
```

#### Saturation analysis

In order to know whether we would gain more from sequencing deeper, we downsampled the data (the final.bam file) to contain 10%, 20%... 90% reads, and then we created the DigitalExpression matrix (as in the normal dropseq pipeline).

This can give us insight, whether we have reached the saturation point (in terms of median umi per cell and median genes per cell) or whether we should sequence deeper.

Results of this are plotted below.

```{r read_summaries, echo = F}
downsampled_summaries <- snakemake@input[startsWith(names(snakemake@input), 'downsampled')]

dge_data <- tibble(percentage = names(downsampled_summaries),
                   filename = unlist(downsampled_summaries)) %>%
            mutate(content = map(filename, 
                                 ~ read_table2(., skip=6))) %>%
            unnest(content) %>%
            select(-filename) %>%
            separate(percentage, into = c(NA, 'percentage')) %>%
            mutate(pcr = NUM_GENIC_READS / NUM_TRANSCRIPTS) %>%
            rename(umis = NUM_TRANSCRIPTS,
                   genes = NUM_GENES,
                   reads = NUM_GENIC_READS)
```
#### Plot distribution of observations

```{r plot_distributions_of_observations, echo =F, fig.width=10}
cPalette = list('grey'= "#999999", 'light_orange'="#E69F00",
         'light_blue'="#56B4E9", 'green' = "#009E73",
         'yellow' = "#F0E442", 'blue'= "#0072B2", 'orange'="#D55E00",
         'pink'="#CC79A7")

clrs = c('umis'=cPalette$light_orange, 'pcr' = cPalette$light_blue, 'reads'=cPalette$green,
         'genes'=cPalette$pink, 'n_beads'= 'black')

median_clrs = clrs
names(median_clrs) = paste0('median_', names(clrs))
clrs = c(clrs, median_clrs)

plot_observations <- function(metric, log_scale = T){
    pl <- dge_data %>%
        select(CELL_BARCODE, percentage, pcr, umis, reads) %>%
        gather('obs', 'val', umis, pcr, reads) %>%
        filter(obs == metric) %>%
        mutate(percentage = as.integer(percentage)) %>%
        group_by(percentage, obs) %>%
        filter(between(val, quantile(val, 0.05, na.rm=T), quantile(val, 0.95, na.rm=T))) %>%
        ggplot(aes(val, fill = obs)) +
            geom_density() +
            facet_grid(percentage~obs, scales = 'free_y') +
            scale_fill_manual(values = clrs) +
            labs(x='') +
            theme(legend.position='none')
    if (log_scale){
        pl <- pl +
            scale_x_log10()+
            annotation_logticks(sides='b')
    }
    return(pl)
}

plot_observations('umis')
plot_observations('reads')
plot_observations('pcr', log_scale=F)
```

##### we bin the data into percentalise by reads per bead

```{r plot_by_deciles, echo = F, fig.height=7, fig.width=7}
decile_dat <- dge_data %>%
    group_by(percentage) %>%
    mutate(cumsum_reads = cumsum(reads),
           decile_limit = sum(reads)/10,
           # put beads into deciles by number of reads
           decile = floor(cumsum_reads / decile_limit) + 1) %>%
    # get top 10 deciles, 11 is an artifact of rounding, last beads
    filter(decile < 11) %>%
    group_by(percentage, decile) %>%
    summarise(median_reads = median(reads),
              median_genes = median(genes),
              median_pcr = median(pcr),
              median_umis = median(umis),
              n_beads = n()) %>%
    gather('observation', 'value', median_reads:n_beads) %>%
    mutate(decile = factor(decile), 
           percentage = as.integer(percentage),
           observation = factor(observation, levels = c(
                      'median_reads', 
                      'median_umis', 
                      'median_genes',
                      'median_pcr', 'n_beads')))

decile_dat %>%
    ggplot(aes(percentage, value, color= decile, fill = decile)) +
        geom_smooth(formula = y~log(x), size = 0.6) +
        geom_point(size=2,  color = 'black', pch=21) + 
        scale_x_continuous(breaks=seq(0, 100, 10)) +
        facet_wrap(~observation, scales = 'free_y', ncol=2) +
        labs(y='', x='downsampling percentage')
```

```{r plot_observations, echo = F}
plot_data <- function(umi_cutoff=1){
    dge_data %>%
        filter(umis > umi_cutoff) %>%
        group_by(percentage) %>%
        summarise(median_reads = median(reads),
                  median_umis = median(umis),
                  median_genes = median(genes),
                  median_pcr = median(pcr),
                  n_beads = n()) %>%
    gather('observation', 'value', median_reads:n_beads) %>% 
    as_tibble() %>%
    mutate(percentage = as.integer(percentage),
           observation = factor(observation, levels = c(
                      'median_reads', 
                      'median_umis', 
                      'median_genes',
                      'median_pcr', 'n_beads'))) %>%
    ggplot(aes(percentage, value, color = observation, fill=observation)) +
        scale_color_manual(values=clrs) +
        scale_fill_manual(values=clrs) +
        geom_smooth(formula = y~log(x), size = 0.6) +
        geom_point(size=2, color = 'black', pch=21) + 
        scale_x_continuous(breaks=seq(0, 100, 10)) +
        facet_wrap(~observation, scales = 'free', ncol=3) +
        labs(y='', x='downsampling percentage', color='', fill='') +
        theme(strip.background=element_blank(), strip.text.x=element_blank()) +
        guides(colour = guide_legend(override.aes = list(size=5))) +
        ggtitle('Median values of downsampled data')
}
```
\newpage

##### With UMI cutoff 1

```{r plot_umi_1, echo = F, fig.height = 5, fig.width = 10}
grid.draw(shift_legend(plot_data(1)))
```

\newpage

##### With UMI cutoff 10

```{r plot_umi_10, echo = F, fig.height = 5, fig.width = 10, fig.align='center'}
grid.draw(shift_legend(plot_data(10)))
```

\newpage

##### With UMI cutoff 50

```{r plot_umi_50, echo = F, fig.height = 5, fig.width = 10, fig.align='center'}
grid.draw(shift_legend(plot_data(50)))
```

\newpage

##### With UMI cutoff 100

```{r plot_umi_100, echo = F, fig.height = 5, fig.width = 10, fig.align='center'}
grid.draw(shift_legend(plot_data(100)))
```

##### With UMI cutoff 500

```{r plot_umi_500, echo = F, fig.height = 5, fig.width = 10, fig.align='center'}
grid.draw(shift_legend(plot_data(500)))
```
